Model Finding for Recursive Functions in SMT\thanks{ This research is partially supported by the Inria technological development action ``Contre-exemples Utilisables par Isabelle et Coq'' (CUIC). }

### 2015-05-07 tex/main.tex
Address some of Cesare's comments.

In addition to standard quantified formulas $\forall \vec x.\; \varphi$,
we consider \emph{annotated quantified formulas} of the form
$\forallf{\const{f}} \vec x.\; \varphi$, where $\con{f} \in \sfuns{\Sigma}$ is
an uninterpreted function symbol, 
whose semantics is the same as standard quantified formulas.
Given $\con{f} : \tau_1 \times \cdots \times \tau_n \rightarrow \tau$,
an annotated
quantified formula $\forallf{\con{f}} \vec x.\; \varphi$ is a \emph{function definition}
(\,\emph{for $\con{f}$}\vthinspace) if $\vec x$ is a tuple of variables
$x_1, \ldots, x_n$ of respective type
$\tau_1, \ldots, \tau_n$ and $\varphi$ is a quantifier-free formula 
$T$-equivalent to $\con{f}( \vec x ) \teq t$ for some term $t$ of type $\tau$.
\ct{Why can't $\varphi$ be existential instead of just quantifier free?} 
We write $\exists
\vec x.\; \varphi$ as a shorthand for $\neg \forall \vec x.\; \neg \varphi$.

@@

\begin{definition}\rm
Given a set of function definitions 
$\DDD = \{ \forallf{\con{f}_1} \vec x.\; \varphi_1, \ldots, \forallf{\con{f}_n} \vec x.\; \varphi_n \}$, 
a formula $\varphi$ 
\ct{I suppose $\varphi$ is meant to be ground.}
is \emph{closed under function expansion 
with respect to $\DDD$} if 
\[
 \varphi \,\models_T\, \bigwedge\nolimits_{\,i=1}^{n} \{ \varphi_i[ \vec t / \vec x ] \mid \con{f}_i( \vec t ) \in \terms( \varphi ) \} .
\]
The set $\DDD$ is \emph{admissible} if for every $T$-satisfiable formula 
$\varphi$ closed under function expansion wrt $\DDD$
the formula
$\varphi \wedge \forallf{\con{f}_1} \vec x.\; \varphi_1 \wedge \cdots \wedge \forallf{\con{f}_n} \vec x.\; \varphi_n$ is also $T$-satisfiable.
\ct{I}
\end{definition}

For example, if $T$ is the theory of integer arithmetic,
the set $\{ \forallf{\con{f}} \vec x:\typeint.\;\, \con{f}( x ) \teq 0 \}$ is admissible.
An admissible set can include definitions of nonterminating functions,
\ct{There is no notion of termination in FOL. We either define this properly or we just replace "definitions of nonterminating functions, e.g.," with "definitions like"}
e.g.,
$\{ \forallf{\con{f}} x : \typeint.\;\, \con{f}( x ) \teq \con{f}( x - 1 ) \}$ or even $\{ \forallf{\con{f}} x.\allowbreak\;\, \con{f}( x ) \teq \con{f}( x ) \}$ are admissible.
On the other hand, the set $\{ \forallf{\con{f}} x.\;\, \con{f}( x ) \teq \con{f}( x ) + 1 \}$ is not admissible,
since the formula $\ltrue$ is (trivially) closed under function expansion with respect to this set,
and there is no model of $T$ satisfying $\forallf{\con{f}} x.\;\, \con{f}( x ) \teq \con{f}( x ) + 1$.
Consider the set 
$\{ 
 \forallf{\con{f}} x.\;\, \con{f}( x ) \teq \con{f}( x ),\;
 \forallf{\con{g}} x.\;\, \lite\bigl( \con{f}( x ) \teq 5,\;
   \con{g}( x ) \teq \con{f}( x ),\,
   \con{g}( x ) \teq \con{g}( x ) + 1)
\}$.
While this set has a model where $\con f$ and $\con g$ are interpreted as the constant $5$, it is not admissible
since the formula $\con{f}( 0 ) \teq 4$ is closed under function expansion with respect to this set,
but there exists no interpretation satisfying both $\con{f}( 0 ) \teq 4$ and $\forallf{\con{g}} x.\;\, \lite( \con{f}( x ) \teq 5,\, \con{g}( x ) \teq \con{f}( x ),\, \con{g}( x ) \teq \con{g}( x ) + 1 )$.

@@

\begin{example}
\label{ex:translation}
Consider the combined signature of linear arithmetic and uninterpreted
functions $\Sigma = \Sigmalia \mathrel{\cup} \Sigma_{\euf}$, where
$\stypes{\Sigma_{\euf}} = \{ \typeint \}$ 
\ct{shouldn't this be $\stypes{\Sigmalia} = \{ \typeint \}$?}
and
$\sfuns{\Sigma_{\euf}} = \{ \con{s} : { \typeint \rightarrow \typeint },\, \const{c} : { \typeint } \}$.
Let $\varphi$ be the $\Sigma$-formula
\begin{equation} \label{eq:ex-before}
\forallf{\con{s}} n : {\typeint}.\;\, \lite\bigl( n \leq 0,\; \con{s}(n) \teq 0,\;
  \con{s}( n ) \teq n + \con{s}( n - 1 ) \bigr) \wedge \con{s}( \con{c} ) > 100
\end{equation}
%
Here, $\const{s}$ is a function that returns the sum of the positive integers
up to $n$. The formula $\varphi$ is in definitional form with
respect to $\sfundefs{\Sigma_u}$ and states that the sum of all
positive numbers between $0$ and $\const{c}$ is greater than $100$. This formula is
satisfiable with a model that interprets $\const{c}$ as an integer greater than or
equal to $14$. Due to the universal quantifier,
current SMT techniques~\cite{GeDeM-CAV-09, ReyEtAl-1-RR-13,reynolds-et-al-2013} are unable to find
a model for this formula. The signature $\extendsig{\Sigma}$ includes the type
$\fargtype{\con{s}}$ and the uninterpreted function $\farg{\con{s}}$ of type $\fargtype{\con{s}}
\rightarrow \typeint$. The result of $\conv( \varphi,\, \ppos )$ is the pair $(
\psi,\, \ltrue )$, where (after simplification) $\psi$ is the
$\extendsig{\Sigma}$-formula
%
\begin{equation} \label{eq:ex-after}
\!\begin{aligned}[c]
  & \phantom{{\wedge}\; }\bigl(
      \forall a : \fargtype{\con{s}}.\; \lite\bigl(
        \!\begin{aligned}[t]
         &  \farg{\con{s}}( a ) \leq 0,\;
          \con{s}(\farg{\con{s}}( a )) \teq 0,\;
\\[-\jot]
  & \con{s}(\farg{\con{s}}( a )) \teq \farg{\con{s}}( a ) + \con{s}( \farg{\con{s}}( a )-1 )
    \wedge \exists b : {\fargtype{\con{s}}}.\;\, \farg{\con{s}}( b ) \teq \farg{\con{s}}( a )-1 \bigr) \bigr)
\end{aligned}
\\[-\jot]
 & {\wedge}\; \bigl( \con{s}( \con{c} ) > 100 \wedge \exists a : {\fargtype{\con{s}}}.\;\, \farg{\con{s}}( a ) \teq \con{c} \bigr)
\end{aligned}
\end{equation}
%
The universal quantification in the resulting formula is over an uninterpreted
type $\fargtype{\con{s}}$, thus enabling the finite model finding
techniques by Reynolds et al.\ \cite{ReyEtAl-1-RR-13,reynolds-et-al-2013},
implemented in \cvc, to search for a finite interpretation for $\fargtype{\con{s}}$. 
Since all occurrences of the quantified variable $a$ in this formula are 
beneath applications of the uninterpreted function $\farg{\con{s}}$, 
the formula is in the essentially uninterpreted fragment,
for which Ge and de Moura \cite{GeDeM-CAV-09} provide 
a complete instantiation procedure, implemented in \ziii. 
Both
%%% Added "apparently". Or are we sure?  
% Yes, I'm planning to post the 2 versions of the benchmark on the evaluation page (when we add it) --ajr
\cvc and \ziii run indefinitely on formula~(\ref{eq:ex-before}), 
while they both produce a model for formula~(\ref{eq:ex-after}) 
within 100 milliseconds.\xend
\end{example}

The translation $\conv$ does not preserve the models of $\varphi$.
\ct{"Preserves models" means that it does not drop models, not that it adds models.}
One model $\I$ for formula~(\ref{eq:ex-after}) in the previous example interprets
$\fargtype{\con{s}}$ as a finite set of elements $\{ u_0, \ldots, u_{14} \}$,
$\farg{\con{s}}$ as a finite map $u_i \mapsto i$ for $i = 0, \ldots, 14$,
$\con{c}$ as $14$,
and $\con{s}$ as the (almost constant) function
%
%% TYPESETTING: \mathbin and to make it fit nicely
\begin{equation} \label{eq:approx-interp}
\lambda x : \typeint.\;\, \lite( x \mathbin{\teq} 0,\, 0,\,
  \lite( x \mathbin{\teq} 1,\, 1,\,
    \lite( x \mathbin{\teq} 2,\, 3,\,
      \lite(\ldots,\; \lite( x \mathbin{\teq} 13,\, 91,\, 105 )\negvthinspace \ldots ))))
\end{equation}
%
In other words, $\const{s}$ is interpreted as a function mapping $x$ to the sum
of all positive integers between $0$ and $x$ when $0 \leq x \leq 13$, and $105$
otherwise.
The interpretation $\I$, restricted to $\Sigma$, is {not} a model for the original formula~(\ref{eq:ex-before}),
since $\I$ wrongly interprets $\con{s}( x )$ as $105$ when $x < 0$ or $x > 14$.

@@

\begin{example}
Let us revisit the translated formula~(\ref{eq:ex-after}) from
Example~\ref{ex:translation}. This formula is unsatisfiable if and only if
formula~(\ref{eq:ex-before}) is unsatisfiable since $\fargtype{f}$ can be
interpreted as $\typeint$, and $\farg{f}$ can be interpreted as the identity
function. Conversely, we claim formula~(\ref{eq:ex-after}) is satisfiable if
and only if formula~(\ref{eq:ex-before}) is satisfiable, noting that the set
$\{ \forallf{\con{s}} x\typ{\typeint}.\;\, \lite\bigl( x \leq 0,\allowbreak\; \con{s}(x)
\teq 0,\; \con{s}( x ) \teq x + \con{s}( x-1 ) \bigr) \}$ is admissible.
\ct{
You cannot possibly mean "if and only if" in the two sentences above.
Once I show that (1) is unsat iff (2) is unsat, I do not need to argue that
(1) is sat iff (2) is sat. That is immediate.
}
As stated
in the proof of Theorem~\ref{thm:equi}, the reason is that any model of
formula~(\ref{eq:ex-after}) must satisfy a formula that is closed under
function expansion, hence implying that formula~(\ref{eq:ex-before}) has a
model. In more detail, let $\varphi_\con{s}$ be the formula $\forallf{\con{s}} x.\;\,
\lite\bigl( x \leq 0, \con{s}(x) \teq 0, \con{s}( x ) \teq x + \con{s}( x-1 )
\bigr)$. The rightmost conjunct $\exists a : {\fargtype{\con{s}}}.
\farg{\con{s}}( a ) \teq \con{c}$ ensures that any model of
formula~(\ref{eq:ex-after}) satisfying $\con{s}( \con{c} ) > 100$ also
satisfies $\varphi_{\con{s}}[\con{c}/x]$. The nested existential
$\exists b.\;\, \farg{\con{s}}( b ) \teq \farg{\con{s}}( a )-1$ ensures that
a similar constraint is enforced recursively: when $\varphi_{\con{s}}[t/x]$ is used in part to evaluate $\con{s}( \con{c} )$,
any model of $\varphi_{\con{s}}[t/x]$ must also satisfy $\varphi_{\con{s}}[t-1/x]$ when $t > 0$. Given a model $\I$ of
formula~(\ref{eq:ex-after}) that interprets $\con{c}$ as $14$, it is easy to see that
due to the structure of formula~(\ref{eq:ex-after}),
$\I$ must satisfy the following formula $\psi$:
%
\begin{equation} \label{eq:fun-closure}
\begin{split}
& \con{s}( \con{c} ) > 100 \wedge %\displaystyle
  \bigwedge\nolimits_{\vthinspace i=0}^{\!13} \bigl( \neg (\con{c}-i \leq 0) \wedge \con{s}( \con{c}-i ) \teq i + \con{s}( \con{c}-i-1 ) \bigr) \\[-\jot]
{\wedge}\; & \con{c}-14 \leq 0 \wedge \con{s}( \con{c}-14 ) \teq 0
\end{split}
\end{equation}
%
This formula is closed under function expansion,
since it contains the $\con{s}$ applications corresponding to
$\con{s}( \con{c}-i )$ for $i = 0, \ldots, 14$. The formula also
entails $\varphi_\con{s}[\con{c}-i/x]$ for $i = 0, \ldots, 14$.
Since $\varphi_\con{s}$ is an admissible definition,
there must exist a $\Sigma$-interpretation satisfying $\psi \wedge \varphi_\con{s}$
which entails formula~(\ref{eq:ex-before}).\xend
\end{example}

@@

\jb{Add link to our evaluation data, like we did in CADE paper}

In this section, we evaluate both the overall impact of the encoding introduced
in the previous section and the performance of individual SMT techniques for
benchmarks in the encoding.

@@

For each of these 245 benchmarks, we considered (up to) three randomly selected
mutated forms of its conjecture $\psi$. In particular, we considered unique
conjectures that are obtained as a result of swapping a subterm of $\psi$ at
one position with another of the same type at another position.
Note that benchmarks created in this way have a high likelihood of having small
and easy-to-find counterexamples \ct{contermodels?}. In total, we considered 213 mutated forms of
conjectures from \isa, and 427 mutated forms of conjectures from \leon. We will
call these sets \isam and \leonm respectively.

@@

For solvers, we considered the SMT solver \ziii~\cite{de-moura-bjoerner-2008},
which runs both heuristic methods for quantifier instantiation~\cite{DBLP:conf/cade/MouraB07},
as well as methods for finding models for quantified formulas~\cite{GeDeM-CAV-09}.
We also considered three configurations of \cvc~\cite{barrett-et-al-2011} which
we will refer to as \cvcd, \cvcf, and \cvci.
The default configuration \cvcd runs heuristic and conflict-based techniques for quantifier instantiation~\cite{ReynoldsTinelliMoura14},
but does not include techniques for finding models.
The configuration \cvcf runs heuristic instantiation and the finite model
finding procedure described in~\cite{ReyEtAl-1-RR-13,reynolds-et-al-2013}. The
last configuration \cvci incorporates techniques for automating inductive
reasoning in SMT~\cite{reynolds-kuncak-2015}.

@@

We described the most closely related work, by Ge and de Moura
\cite{GeDeM-CAV-09} and by Reynolds et al.\
\cite{ReyEtAl-1-RR-13,reynolds-et-al-2013}, in the text already.
Outside the world of SMT, model finding has been
studied many times. 
The text below is partly adapted from Blanchette's Ph.D.\
thesis \cite{blanchette-2012-phd}.
\ct{The last sentence is strange. I would remove it since Jasmin is one of the authors.}

@@

The Leon system \cite{suter-et-al-2011} implements a procedure that can both
produces proofs and counterexamples for properties of terminating functions
written in a subset of Scala. Leon is based on an SMT solver. It avoids
quantifiers altogether by unfolding recursive definitions up to a certain
depth, which is increased iteratively as necessary.
\jb{Andy: Can you add some more here? You're the Leon pro! E.g. why/when do we think
that our approach might be better?}

\ct{About Leon, the reviewers and readers may wonder why we do not do the same, that is, just progressively unfold the recursive definitions until a model is found.
We should probably say something in this regard.
}

>>>>

In addition to standard quantified formulas $\forall \vec x.\; \varphi$,
we consider \emph{annotated quantified formulas} of the form
$\forallf{\const{f}} \vec x.\; \varphi$, where $\con{f} \in \sfuns{\Sigma}$ is
an uninterpreted function symbol, 
whose semantics is the same as standard quantified formulas.
Given $\con{f} : \tau_1 \times \cdots \times \tau_n \rightarrow \tau$,
an annotated
quantified formula $\forallf{\con{f}} \vec x.\; \varphi$ is a \emph{function definition}
(\,\emph{for $\con{f}$}\vthinspace) if $\vec x$ is a tuple of variables
$x_1, \ldots, x_n$ of respective type
$\tau_1, \ldots, \tau_n$ and $\varphi$ is a quantifier-free formula 
$T$-equivalent to $\con{f}( \vec x ) \teq t$ for some term $t$ of type $\tau$.
%\ct{Why can't $\varphi$ be existential instead of just quantifier free?} 
%I don't think it makes much sense to have any quantifiers within function bodies, 
%since they aren't operational, that's why I put quantifier-free. --ajr
We write $\exists
\vec x.\; \varphi$ as a shorthand for $\neg \forall \vec x.\; \neg \varphi$.

@@

\begin{definition}\rm
Given a set of function definitions 
$\DDD = \{ \forallf{\con{f}_1} \vec x.\; \varphi_1, \ldots, \forallf{\con{f}_n} \vec x.\; \varphi_n \}$, 
a ground formula $\varphi$ 
%\ct{I suppose $\varphi$ is meant to be ground.} yes --ajr
is \emph{closed under function expansion with respect to $\DDD$} if 
\[
 \varphi \,\models_T\, \bigwedge\nolimits_{\,i=1}^{n} \{ \varphi_i[ \vec t / \vec x ] \mid \con{f}_i( \vec t ) \in \terms( \varphi ) \} .
\]
The set $\DDD$ is \emph{admissible} if for every $T$-satisfiable formula 
$\varphi$ closed under function expansion wrt $\DDD$
the formula
$\varphi \wedge \forallf{\con{f}_1} \vec x.\; \varphi_1 \wedge \cdots \wedge \forallf{\con{f}_n} \vec x.\; \varphi_n$ is also $T$-satisfiable.
\ct{I}
\end{definition}

For example, if $T$ is the theory of integer arithmetic,
the set $\{ \forallf{\con{f}} \vec x:\typeint.\;\, \con{f}( x ) \teq 0 \}$ is admissible.
An admissible set can include definitions of nonterminating functions,
\ct{There is no notion of termination in FOL. We either define this properly or we just replace "definitions of nonterminating functions, e.g.," with "definitions like"}
\ajr{I agree. Jasmin, can you fix this?  We need to be careful in the same way when we talk about the "terminating" functions fragment we handle.}
e.g.,
$\{ \forallf{\con{f}} x : \typeint.\;\, \con{f}( x ) \teq \con{f}( x - 1 ) \}$ or even $\{ \forallf{\con{f}} x.\allowbreak\;\, \con{f}( x ) \teq \con{f}( x ) \}$ are admissible.
On the other hand, the set $\{ \forallf{\con{f}} x.\;\, \con{f}( x ) \teq \con{f}( x ) + 1 \}$ is not admissible,
since the formula $\ltrue$ is (trivially) closed under function expansion with respect to this set,
and there is no model of $T$ satisfying $\forallf{\con{f}} x.\;\, \con{f}( x ) \teq \con{f}( x ) + 1$.
Consider the set 
$\{ 
 \forallf{\con{f}} x.\;\, \con{f}( x ) \teq \con{f}( x ),\;
 \forallf{\con{g}} x.\;\, \lite\bigl( \con{f}( x ) \teq 5,\;
   \con{g}( x ) \teq \con{f}( x ),\,
   \con{g}( x ) \teq \con{g}( x ) + 1)
\}$.
While this set has a model where $\con f$ and $\con g$ are interpreted as the constant $5$, it is not admissible
since the formula $\con{f}( 0 ) \teq 4$ is closed under function expansion with respect to this set,
but there exists no interpretation satisfying both $\con{f}( 0 ) \teq 4$ and $\forallf{\con{g}} x.\;\, \lite( \con{f}( x ) \teq 5,\, \con{g}( x ) \teq \con{f}( x ),\, \con{g}( x ) \teq \con{g}( x ) + 1 )$.

@@

\begin{example}
\label{ex:translation}
Consider the combined signature of linear arithmetic and uninterpreted
functions $\Sigma = \Sigmalia \mathrel{\cup} \Sigma_{\euf}$, where
%$\stypes{\Sigma_{\euf}} = \{ \typeint \}$ 
%\ct{shouldn't this be $\stypes{\Sigmalia} = \{ \typeint \}$?}
%and
% Something went wrong, let's just remove it --ajr
$\sfuns{\Sigma_{\euf}} = \{ \con{s} : { \typeint \rightarrow \typeint },\, \const{c} : { \typeint } \}$.
Let $\varphi$ be the $\Sigma$-formula
\begin{equation} \label{eq:ex-before}
\forallf{\con{s}} n : {\typeint}.\;\, \lite\bigl( n \leq 0,\; \con{s}(n) \teq 0,\;
  \con{s}( n ) \teq n + \con{s}( n - 1 ) \bigr) \wedge \con{s}( \con{c} ) > 100
\end{equation}
%
Here, $\const{s}$ is a function that returns the sum of the positive integers
up to $n$. The formula $\varphi$ is in definitional form with
respect to $\sfundefs{\Sigma_u}$ and states that the sum of all
positive numbers between $0$ and $\const{c}$ is greater than $100$. This formula is
satisfiable with a model that interprets $\const{c}$ as an integer greater than or
equal to $14$. Due to the universal quantifier,
current SMT techniques~\cite{GeDeM-CAV-09, ReyEtAl-1-RR-13,reynolds-et-al-2013} are unable to find
a model for this formula. The signature $\extendsig{\Sigma}$ includes the type
$\fargtype{\con{s}}$ and the uninterpreted function $\farg{\con{s}}$ of type $\fargtype{\con{s}}
\rightarrow \typeint$. The result of $\conv( \varphi,\, \ppos )$ is the pair $(
\psi,\, \ltrue )$, where (after simplification) $\psi$ is the
$\extendsig{\Sigma}$-formula
%
\begin{equation} \label{eq:ex-after}
\!\begin{aligned}[c]
  & \phantom{{\wedge}\; }\bigl(
      \forall a : \fargtype{\con{s}}.\; \lite\bigl(
        \!\begin{aligned}[t]
         &  \farg{\con{s}}( a ) \leq 0,\;
          \con{s}(\farg{\con{s}}( a )) \teq 0,\;
\\[-\jot]
  & \con{s}(\farg{\con{s}}( a )) \teq \farg{\con{s}}( a ) + \con{s}( \farg{\con{s}}( a )-1 )
    \wedge \exists b : {\fargtype{\con{s}}}.\;\, \farg{\con{s}}( b ) \teq \farg{\con{s}}( a )-1 \bigr) \bigr)
\end{aligned}
\\[-\jot]
 & {\wedge}\; \bigl( \con{s}( \con{c} ) > 100 \wedge \exists a : {\fargtype{\con{s}}}.\;\, \farg{\con{s}}( a ) \teq \con{c} \bigr)
\end{aligned}
\end{equation}
%
The universal quantification in the resulting formula is over an uninterpreted
type $\fargtype{\con{s}}$, thus enabling the finite model finding
techniques by Reynolds et al.\ \cite{ReyEtAl-1-RR-13,reynolds-et-al-2013},
implemented in \cvc, to search for a finite interpretation for $\fargtype{\con{s}}$. 
Since all occurrences of the quantified variable $a$ in this formula are 
beneath applications of the uninterpreted function $\farg{\con{s}}$, 
the formula is in the essentially uninterpreted fragment,
for which Ge and de Moura \cite{GeDeM-CAV-09} provide 
a complete instantiation procedure, implemented in \ziii. 
Both
%%% Added "apparently". Or are we sure?  
% Yes, I'm planning to post the 2 versions of the benchmark on the evaluation page (when we add it) --ajr
\cvc and \ziii run indefinitely on formula~(\ref{eq:ex-before}), 
while they both produce a model for formula~(\ref{eq:ex-after}) 
within 100 milliseconds.\xend
\end{example}

The translation $\conv$ results in formulas whose models are different than those of $\varphi$.
%\ct{"Preserves models" means that it does not drop models, not that it adds models.}
One model $\I$ for formula~(\ref{eq:ex-after}) in the previous example interprets
$\fargtype{\con{s}}$ as a finite set of elements $\{ u_0, \ldots, u_{14} \}$,
$\farg{\con{s}}$ as a finite map $u_i \mapsto i$ for $i = 0, \ldots, 14$,
$\con{c}$ as $14$,
and $\con{s}$ as the (almost constant) function
%
%% TYPESETTING: \mathbin and to make it fit nicely
\begin{equation} \label{eq:approx-interp}
\lambda x : \typeint.\;\, \lite( x \mathbin{\teq} 0,\, 0,\,
  \lite( x \mathbin{\teq} 1,\, 1,\,
    \lite( x \mathbin{\teq} 2,\, 3,\,
      \lite(\ldots,\; \lite( x \mathbin{\teq} 13,\, 91,\, 105 )\negvthinspace \ldots ))))
\end{equation}
%
In other words, $\const{s}$ is interpreted as a function mapping $x$ to the sum
of all positive integers between $0$ and $x$ when $0 \leq x \leq 13$, and $105$
otherwise.
The interpretation $\I$, restricted to $\Sigma$, is {not} a model for the original formula~(\ref{eq:ex-before}),
since $\I$ wrongly interprets $\con{s}( x )$ as $105$ when $x < 0$ or $x > 14$.

@@

\begin{example}
Let us revisit the translated formula~(\ref{eq:ex-after}) from
Example~\ref{ex:translation}. This formula is unsatisfiable only if
formula~(\ref{eq:ex-before}) is unsatisfiable since $\fargtype{f}$ can be
interpreted as $\typeint$, and $\farg{f}$ can be interpreted as the identity
function. Conversely, we claim formula~(\ref{eq:ex-after}) is satisfiable
only if formula~(\ref{eq:ex-before}) is satisfiable, noting that the set
$\{ \forallf{\con{s}} x\typ{\typeint}.\;\, \lite\bigl( x \leq 0,\allowbreak\; \con{s}(x)
\teq 0,\; \con{s}( x ) \teq x + \con{s}( x-1 ) \bigr) \}$ is admissible.
%\ct{
%You cannot possibly mean "if and only if" in the two sentences above.
%Once I show that (1) is unsat iff (2) is unsat, I do not need to argue that
%(1) is sat iff (2) is sat. That is immediate.
%}
% You're right, changed both to "only if" --ajr
As stated
in the proof of Theorem~\ref{thm:equi}, the reason is that any model of
formula~(\ref{eq:ex-after}) must satisfy a formula that is closed under
function expansion, hence implying that formula~(\ref{eq:ex-before}) has a
model. In more detail, let $\varphi_\con{s}$ be the formula $\forallf{\con{s}} x.\;\,
\lite\bigl( x \leq 0, \con{s}(x) \teq 0, \con{s}( x ) \teq x + \con{s}( x-1 )
\bigr)$. The rightmost conjunct $\exists a : {\fargtype{\con{s}}}.
\farg{\con{s}}( a ) \teq \con{c}$ ensures that any model of
formula~(\ref{eq:ex-after}) satisfying $\con{s}( \con{c} ) > 100$ also
satisfies $\varphi_{\con{s}}[\con{c}/x]$. The nested existential
$\exists b.\;\, \farg{\con{s}}( b ) \teq \farg{\con{s}}( a )-1$ ensures that
a similar constraint is enforced recursively: when $\varphi_{\con{s}}[t/x]$ is used in part to evaluate $\con{s}( \con{c} )$,
any model of $\varphi_{\con{s}}[t/x]$ must also satisfy $\varphi_{\con{s}}[t-1/x]$ when $t > 0$. Given a model $\I$ of
formula~(\ref{eq:ex-after}) that interprets $\con{c}$ as $14$, it is easy to see that
due to the structure of formula~(\ref{eq:ex-after}),
$\I$ must satisfy the following formula $\psi$:
%
\begin{equation} \label{eq:fun-closure}
\begin{split}
& \con{s}( \con{c} ) > 100 \wedge %\displaystyle
  \bigwedge\nolimits_{\vthinspace i=0}^{\!13} \bigl( \neg (\con{c}-i \leq 0) \wedge \con{s}( \con{c}-i ) \teq i + \con{s}( \con{c}-i-1 ) \bigr) \\[-\jot]
{\wedge}\; & \con{c}-14 \leq 0 \wedge \con{s}( \con{c}-14 ) \teq 0
\end{split}
\end{equation}
%
This formula is closed under function expansion,
since it contains the $\con{s}$ applications corresponding to
$\con{s}( \con{c}-i )$ for $i = 0, \ldots, 14$. The formula also
entails $\varphi_\con{s}[\con{c}-i/x]$ for $i = 0, \ldots, 14$.
Since $\varphi_\con{s}$ is an admissible definition,
there must exist a $\Sigma$-interpretation satisfying $\psi \wedge \varphi_\con{s}$
which entails formula~(\ref{eq:ex-before}).\xend
\end{example}

@@

In this section, we evaluate both the overall impact of the encoding introduced
in the previous section and the performance of individual SMT techniques for
benchmarks in the encoding.
The experimental data is publicly available.%
\footnote{\url{http://lara.epfl.ch/~reynolds/SMT2015-recfun/}}

@@

For each of these 245 benchmarks, we considered (up to) three randomly selected
mutated forms of its conjecture $\psi$. In particular, we considered unique
conjectures that are obtained as a result of swapping a subterm of $\psi$ at
one position with another of the same type at another position.
Note that benchmarks created in this way have a high likelihood of having small
and easy-to-find countermodels. In total, we considered 213 mutated forms of
conjectures from \isa, and 427 mutated forms of conjectures from \leon. We will
call these sets \isam and \leonm respectively.

@@

For solvers, we considered the SMT solver \ziii~\cite{de-moura-bjoerner-2008},
which runs both heuristic methods for quantifier instantiation~\cite{MouraBjoerner07},
as well as methods for finding models for quantified formulas~\cite{GeDeM-CAV-09}.
We also considered three configurations of \cvc~\cite{barrett-et-al-2011} which
we will refer to as \cvcd, \cvcf, and \cvci.
The default configuration \cvcd runs heuristic and conflict-based techniques for quantifier instantiation~\cite{ReynoldsTinelliMoura14},
but does not include techniques for finding models.
The configuration \cvcf runs heuristic instantiation and the finite model
finding procedure described in~\cite{ReyEtAl-1-RR-13,reynolds-et-al-2013}. The
last configuration \cvci incorporates techniques for automating inductive
reasoning in SMT~\cite{reynolds-kuncak-2015}.

@@

We described the most closely related work, by Ge and de Moura
\cite{GeDeM-CAV-09} and by Reynolds et al.\
\cite{ReyEtAl-1-RR-13,reynolds-et-al-2013}, in the text already.
Outside the world of SMT, model finding has been
studied many times.

@@

The Leon system \cite{suter-et-al-2011} implements a procedure that can both
produces proofs and counterexamples for properties of terminating functions
written in a subset of Scala. Leon is based on an SMT solver. It avoids
quantifiers altogether by unfolding recursive definitions up to a certain
depth, which is increased iteratively as necessary.
\jb{Andy: Can you add some more here? You're the Leon pro! E.g. why/when do we think
that our approach might be better?}
\ct{About Leon, the reviewers and readers may wonder why we do not do the same, that is, just progressively unfold the recursive definitions until a model is found.
We should probably say something in this regard.
}
